### 2017-02-08 doc/manuscript.Rmd
Edits as per Tony's suggestions from email (esp re: vs Tg)

Hypertriglyceridemia is an extensively studied and well known factor involved in
the dysregulation of metabolic function and subsequent negative health outcomes {{cite}}.
It is a component of the metabolic syndrome [@Alberti2009], a strong risk factor for
cardiovascular disease [@DAgostino2004a], and has been shown to contribute to the pathogenesis
of diabetes {{cite}}. The measurement of circulating triacylglcerides (TAG) is commonly done
during routine clinical visits as part of health assessment. However, clinically
measured TAG is limited as it represents all the types of fatty acids within the
TAG fraction as a single measure. Given the increasing appreciation for the role
of the specific fatty acid composition on health outcomes, for instance with
differences in associations of fatty acids within the PL lipid fraction
[@Forouhi2014a; @Ma2015a] including an analysis from the PROMISE cohort
[@Johnston2016a], there are relatively few studies that have explored this area
of research in the TAG fraction.

The interaction between TAG and insulin sensitivity is complex and has 
components of a feedback system {{cite}}. Greater resistance to insulin in both the liver
and muscle may result in greater production of TAG and secretion of lipoproteins
that transport TAG [@Yu2012a]. Likewise, greater TAGFA may contribute to 
metabolic dysfunction and lipotoxicity in various tissues, such as the 
beta-cells, and thus continuing the cycle {{cite}}. Given the complexity and temporal
nature of the relationship, long term studies with multiple data collections are
paramount to better understanding the underlying biology and subsequent risk.

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S
[@Levy1998a] using the HOMA2 Calculator and the Insulin Sensitivity Index (ISI)
[@Matsuda1999]. HOMA largely reflects hepatic insulin resistance, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007]. Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR [@Matthews1985] (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
the disposition index (but using OGTT values). Each index has been validated
against gold standard measures [@Matthews1985; @Hermans1999a; @Matsuda1999; @Retnakaran2009].
Specific formulas for each OGTT-derived measure can be found in the {{
Supplementary Material }}.

@@

Previous research has shown that carbohydrate intake increases DNL 
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study
extends these findings by showing that TAGFA with 14 to 16 carbons clustered together and all
strongly predicted lower insulin sensitivity. While these fatty acids also had a
significant association with beta-cell function, the results were not as strong
as with the insulin sensitivity.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A small, short feeding trial (n=24) was conducted to identify
fatty acids that most accurately reflected DNL, as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TAGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together using the PLS analysis, implicating these fatty acids as
indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
particularly one characterized by a predominance of simple carbohydrates, may
lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
fat stores and consequently increasing the amount of TAGFA in circulation {{cite}}. The higher
concentration of circulating 14 and 16 carbon fatty acids may then expose
tissues to greater lipotoxicity, for instance from palmitic acid (16:0), which
is well-known to have harmful effects on tissues [@Riserus2008a; @Iggman2010a].

We found no interaction by time for any of the TAGFA on the metabolic outcomes.
Analyzing these biological processes longitudinally is challenging due to the
potential feedback mechanisms involved. For instance, while greater DNL may
promote muscle insulin resistance, the reverse may also be true
[@Flannery2012a]. Higher insulin resistance may encourage greater DNL to handle
the higher blood glucose. To illustrate this, in a weight loss intervention trial
(n=19 with TAGFA data), participants who lost weight over 33 weeks showed higher
insulin sensitivity and lower TAGFA composition indicative of lower DNL (less
14:0, 14:1, 16:0, 16:1, etc) [@Schwab2008a]. Given the complex biological
mechanisms and feedback loops involved, disentangling whether insulin
sensitivity influences TAGFA more strongly than TAGFA influencing insulin
sensitivity will require more complicated research designs and analyses. Our own
null results on this question may tentatively point to an equally influential
feedback mechanism between TAGFA and insulin sensitivity {{rephrase after exploring time interaction}}.

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
from the adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synthesized
VLDL-TAG during fasting and ~40% during the fed state while dietary fatty acids
contributed ~25% to newly synthesized VLDL-TAG [@Barrows2006a]. In an analysis of the
serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
(under review), we found that higher total NEFA but not the specific
composition associated with lower beta-cell function and not with insulin 
resistance. While NEFA may contribute substantially to the TAG fraction, there 
may be something unique about the specific composition of the TAGFA fraction 
that may allow it to enter insulin sensitive tissues more easily than NEFA. It 
may be that VLDL receptors allow fatty acids to enter these tissue more easily
than NEFA bound to albumin, or it may be due to the other influences on TAGFA 
composition such as dietary intake {{cite?}}.

@@

The authors had the following responsibility: LWJ conducted research, analyzed
data, and wrote the paper; RR, ZL, and SBH designed research, conducted
research, and provided essential materials (infrastructure and clinical
resources); RR, SBH, RPB, and AG provided intellectual feedback on the
paper; RPB conducted research, provided essential reagents and materials; AJH
designed research, assisted with interpretation, and provided intellectual
feedback on all versions of the paper; LWJ and AJH had primary responsibility
for final content. All authors read and approved the final manuscript.

{{Bernie?}}

>>>>

Hypertriglyceridemia is an extensively studied and well known factor involved in
the dysregulation of metabolic function and subsequent negative health outcomes [@Chehade2013a].
It is a component of the metabolic syndrome [@Alberti2009] and a strong risk factor for
cardiovascular disease [@DAgostino2004a], in addition to it's role in metabolic
dysfunction [@Verges2015a; @Kawano2013a]. The measurement of circulating 
triacylglcerides (TAG) is commonly done during routine clinical visits as part
of health assessment. However, clinically measured TAG is limited as it
represents all the types of fatty acids within the TAG fraction as a single
measure. Given the increasing appreciation for the role of the specific fatty
acid composition on health outcomes, for instance with differences in
associations of fatty acids within the PL lipid fraction [@Forouhi2014a; @Ma2015a] 
including an analysis from the PROMISE cohort [@Johnston2016a], there
are relatively few studies that have explored this area of research in the TAG
fraction.

The interaction between TAG and insulin sensitivity is complex and has 
components of a feedback system [@Verges2015a]. Greater resistance to insulin in both the liver
and muscle may result in greater production of TAG and secretion of lipoproteins
that transport TAG [@Yu2012a]. Likewise, greater TAGFA may contribute to 
metabolic dysfunction and lipotoxicity in various tissues, such as the 
beta-cells, and thus continuing the cycle [@Verges2015a]. Given the complexity and temporal
nature of the relationship, long term studies with multiple data collections are
paramount to better understanding the underlying biology and subsequent risk.

@@

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S
[@Levy1998a] using the HOMA2 Calculator and the Insulin Sensitivity Index (ISI)
[@Matsuda1999]. HOMA largely reflects hepatic insulin resistance, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007]. Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR [@Matthews1985] (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
the disposition index (but using OGTT values). Each index has been validated
against gold standard measures [@Matthews1985; @Hermans1999a; @Matsuda1999; @Retnakaran2009].
Specific formulas for each OGTT-derived measure can be found in the 
Supplementary Material.

@@

Previous research has shown that carbohydrate intake increases DNL 
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study
extends these findings by showing that TAGFA with 14 to 16 carbons clustered together and all
strongly predicted lower insulin sensitivity. While these fatty acids also had a
significant association with beta-cell function, the results were not as strong
as with the insulin sensitivity. However, while the DNL cluster had a strong
harmful association with insulin sensitivity when we include it in the GEE
models, the magnitude of the association was not as large as when using
clinically measured TAG, reinforcing that the total TAG concentration still more
strongly predicts metabolic dysfunction.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A small, short feeding trial (n=24) was conducted to identify
fatty acids that most accurately reflected DNL, as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TAGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together using the PLS analysis, implicating these fatty acids as
indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
particularly one characterized by a predominance of simple carbohydrates, may
lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
fat stores and consequently increasing the amount of TAGFA in circulation
[@Chehade2013a; @Verges2015a]. The higher concentration of circulating 14 and 16
carbon fatty acids may then expose tissues to greater lipotoxicity, for instance
from palmitic acid (16:0), which is well-known to have harmful effects on
tissues [@Riserus2008a; @Iggman2010a].

We found no interaction by time for any of the TAGFA on the metabolic outcomes.
Analyzing these biological processes longitudinally is challenging due to the
potential feedback mechanisms involved. For instance, while greater DNL may
promote muscle insulin resistance, the reverse may also be true
[@Flannery2012a]. Higher insulin resistance may encourage greater DNL to handle
the higher blood glucose. To illustrate this, in a weight loss intervention trial
(n=19 with TAGFA data), participants who lost weight over 33 weeks showed higher
insulin sensitivity and lower TAGFA composition indicative of lower DNL (less
14:0, 14:1, 16:0, 16:1, etc) [@Schwab2008a]. Given the complex biological
mechanisms and feedback loops involved, disentangling whether insulin
sensitivity influences TAGFA more strongly than TAGFA influencing insulin
sensitivity will require more complicated research designs and analyses. Our own
findings of no interaction by time with the TAGFA tentatively suggest that the
influence of either TAGFA or insulin resistance within the feedback loop may be
of relatively equal strength.

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
from the adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synthesized
VLDL-TAG during fasting and ~40% during the fed state while dietary fatty acids
contributed ~25% to newly synthesized VLDL-TAG [@Barrows2006a]. In an analysis of the
serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
(under review), we found that higher total NEFA but not the specific
composition associated with lower beta-cell function and not with insulin 
resistance. While NEFA may contribute substantially to the TAG fraction, there 
may be something unique about the specific composition of the TAGFA fraction 
that may allow it to enter insulin sensitive tissues more easily than NEFA. It 
may be that VLDL receptors allow fatty acids to enter these tissue more easily
than NEFA bound to albumin or it may be due to the other influences on TAGFA 
composition such as dietary intake since circulating NEFA primarily come from
adipose tissue [@Verges2015a].

@@

The authors had the following responsibility: LWJ conducted research, analyzed
data, and wrote the paper; RR, BZ, ZL, and SBH designed research, conducted
research, and provided essential materials (infrastructure and clinical
resources); RR, BZ, SBH, RPB, and AG provided intellectual feedback on the
paper; RPB and ZL conducted research, provided essential reagents and materials; AJH
designed research, assisted with interpretation, and provided intellectual
feedback on all versions of the paper; LWJ and AJH had primary responsibility
for final content. All authors read and approved the final manuscript.
